Formal Grammars as Types in Non-commutative Bunched Type Theory

### 2023-12-12 paper/paper.tex
Kleene categories

@@

>>>>

\newcommand{\cat}[1]{\mathbf{#1}}

@@

\section{Kleene Category}

An important tool from the theory of formal languages was the development
of Kleene algebras, which serve as a theoretical substrate to studying
various kinds of formal languages. Formally, they are a set equipped
with three operations and two constants satisfying the axioms of
\ref{fig:axioms}.

\begin{figure}
  \begin{align*}
    x + (y + z) &= (x + y) + z & x + y &= y + x\\
    x + 0 &= x & x + x &= x\\
    x(yz) &= (xy)z & x1 &= 1x = x\\
    x(y + z) &= xy + xz & (x + y)z &= xz + yz\\
    x0 &= 0x = x & & \\
    1 + aa^* &\leq a^* & 1 + a^*a &\leq a^*\\
     b + ax \leq x &\implies a^*b \leq x &  b + xa \leq x &\implies ba^* \leq x
  \end{align*}
  \label{fig:axioms}
  \caption{Kleene algebra axioms}
\end{figure}

The addition operation can be used to define a partial order structure
$a \leq b$ if $a + b = b$. In the theory of formal languages this
order structure can be used to model language containment. What we
want to do in this section is to categorify the concept of Kleene
algebra and build on top of this new categorical concept as the
theoretical basis to the formal theory of parsing.

\begin{definition}
  A Kleene category is a distributive monoidal category $\cat{K}$
  such that for every objects $A$ and $B$, the endofunctors $F_{A, B}
  = B + A \otimes X$ and $G_{A, B} = B + X \otimes A$ have initial
  algebras (denoted $\mu X.\, F_{A, B}(X)$) such that $B \otimes (\mu
  X.\, F_{A, 1}) \cong \mu X.\, F_{A, B}(X)$ and the analogue isomorphism
  for $G_{A,B}$ also holds.
\end{definition}

\begin{example}
  Every Kleene algebra, seen a posetal category, is a Kleene category.
\end{example}

\begin{example}
  The opposite category of every Kleene category is a model of a variant of
  conjunctive ordered logic, where the Kleene star plays the role of the ``of
  course'' modality from substructural logics.
\end{example}

The proposed axioms are a direct translation of the Kleene algebra
axioms to a categorical setting. Its most unusual aspect is the
axiomatization of the Kleene star as a family os initial algebras
satisfying certain isomorphisms. If the Kleene category $\cat{K}$ has
more structure, then these isomorphisms hold ``for free''.

\begin{theorem}
  \label{th:kleeneclosed}
  Let $\cat{K}$ be a Kleene category such that it is also monoidal
  closed.  Then, the initial algebras isomorphisms hold automatically.
\end{theorem}
\begin{proof}
  We prove this by the unicity (up-to isomorphism) of initial
  algebras. Let $[hd, tl]: 1 + (\mu X.\, F_{A, 1}(X)) \otimes A \to
  (\mu X.\, F_{A, 1}(X))$ be the initial algebra structure of $(\mu
  X.\, F_{A, 1}(X))$ and consider the map $[hd, tl] : B + B \otimes
  (\mu X.\, F_{A, 1}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
    1}(X))$.

  Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
  we want to show that there is a unique algebra morphism $h : B
  \otimes \mu X.\, F_{A,1} \to Y$. We can show existence and uniqueness
  by showing that the diagram on the left commute if, and only if,
  the diagram on the right commutes:

  This equivalence follows by using the adjunction structure given
  by the monoidal closed structure of $\cat{K}$. A completely analogous
  argument for $G_{A,B}$ also holds.
\end{proof}

This result feels similar in spirit to the definition of action
algebras, which are algebras where the product also has adjoint
operations which results in the Kleene star being more easily
axiomatized \cite{kozen1994}.

%% \begin{theorem}
%%   Let $\cat{K}$ be a Kleene category and $A$ a discrete category.
%%   The functor category $[A, \cat{K}]$.
%%   (HOW GENERAL SHOULD THIS THEOREM BE? BY ASSUMING ENOUGH STRUCTURE,
%%   E.G. K = Set, THIS THEOREM BECOMES SIMPLE TO PROVE)
%% \end{theorem}

\begin{theorem}
  If $\cat{C}$ is a locally small monoidal category, then the category
  $\cat{Set}^{\cat{C}}$ is a Kleene category.
\end{theorem}
\begin{proof}
  The coproduct structure exists since it is a presheaf category. The
  monoidal structure is defined using the Day convolution and,
  furthermore, this monoidal structure has a closed structure as well.

  From this closed structure it follows that it is a distributive
  monoidal category, since left adjoint functors preserve colimits. As
  for the Kleene star, since presheaf categories admit small colimits,
  the initial algebra of the functors $F_{A,B}$ and $G_{A,B}$ can be
  defined as the filtered colimit of the diagrams:

  From Theorem~\ref{th:kleeneclosed} it follows that
  these initial algebras satisfy the required isomorphisms.
\end{proof}

\begin{corollary}
  For every alphabet $\Sigma$, the presheaf category $\cat{Set}^{\cat{\Sigma^*}}$
  is a Kleene category.
\end{corollary}
\begin{proof}
  Note that string concatenation makes the discrete category
  $\Sigma^*$ a strict monoidal category.
\end{proof}
